---
title: "Susanna_Proteomics_follow_up"
author: "Dillon Corvino"
date: "27/01/2025"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: united
    highlight: tango
    df_print: paged
    code_folding: hide
editor_options: 
  chunk_output_type: console
---

Built with R version `r getRversion()`

## Setup {.tabset}

### Dataset information
```{r Dataset_Info}

# Data downloaded from 
# https://zenodo.org/record/8275845


```

### Environment
```{r setup}

knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,  
  eval = TRUE, 
  tidy = TRUE
)

knitr::opts_knit$set(
  root.dir = "../"
)

# Environment Set up
rm(list = ls()) #Clean workspace
cat("\014")     #Clean Console
gc() # Free memory

# Establish working directory
library("rstudioapi")

# Set working directory to source file location
setwd(dirname(getActiveDocumentContext()$path))
setwd("..")

# pipeline variables
long.compute <- FALSE
long.load <- TRUE

# load packages 
library(SeuratDisk)
library(Seurat)
library(scCustomize)
library(dplyr)
library(ggplot2)
library(stringr)


########################################
# Parallelization of seurat functions
########################################

parallelization <- FALSE

if(parallelization){
  library(future)
  # check the current active plan
  plan()
  
  # detect cores
  ncores <- parallel::detectCores()
  ncores <- ncores-5
  # set plan
  plan("multisession", workers = ncores)
  plan()
  
  # set max limit for global variables
  # note this will drastically increase RAM usage, be aware of machine limits
  # set value is in bites therefore 1000*1024^2 = 1GB Therefore set to 10GBs 
  options(future.globals.maxSize = 10000 * 1024^2, 
          future.rng.onMisuse="ignore") 
}



# Create output dir
output.dir <- "results/susanna_proteomics/"

if(!dir.exists(paste0(output.dir))){
  dir.create(paste0(output.dir), 
             recursive = T)
}



# colour scheme 
batlow.pal <- usefulfunctions::Get.batlow()


```

## Load data {.tabset}


### Read seurat files

```{r read_data}

if(long.load){
  
  # unsure what is contained within .*_blood.h5ad object, if this is just a subset or unique data
  # regardless, will just use the fuller object for analysis
  
  # list files
  #files.input <- list.files("data", pattern = "h5seurat", full.names = T)
  
  # Load files
  #seurat.list <- lapply(files.input, LoadH5Seurat)
  #names(seurat.list) <- c("object_1", "object_2")
  
  nk.seurat <- LoadH5Seurat("data/comb_CD56_CD16_NK.h5seurat")
  
  # format metadata 
  nk.seurat@meta.data <- nk.seurat@meta.data %>%
    mutate(tissue = ifelse(meta_tissue %in% c("Normal", "Others"), "non_tumor", as.character(meta_tissue)))
  
}

```


### Read in signature list
```{r read_in_signature}

DE.in.inhibitor.only <- read.delim("data/susanna_signatures/DE_in_inhibitor_only.txt", header = T)
DE.in.inhibitor.only <- DE.in.inhibitor.only$gene.symbol
head(DE.in.inhibitor.only)

DE.in.vehicle.only <- read.delim("data/susanna_signatures/DE_in_vehicle_only.txt", header = T)
DE.in.vehicle.only <- DE.in.vehicle.only$gene.symbol
head(DE.in.vehicle.only)


# Define input and output paths
input.path  <- "data/susanna_signatures/"
output.dir  <- "results/susanna_proteomics_goi/"
dir.create(output.dir, showWarnings = FALSE)

# Read gene lists into named list
gene.list.files <- list.files(input.path, pattern = "\\.txt$", full.names = TRUE)

gene.sets <- lapply(gene.list.files, function(f){
  df <- read.table(f, header = TRUE, sep = "\t", stringsAsFactors = FALSE)
  return(as.character(df[[1]]))  # grab first column regardless of name
})

names(gene.sets) <- basename(gene.list.files) |>
  str_replace_all(" ", "_") |>
  str_remove("\\.txt$")

gene.sets

```

## Dataset visualisation {.tabset}


### visualise genes of interest
```{r vis_genes_of_interest}


goi <- c("DGAT1", "DGAT2", "TMX1", "TMEM68")



# plot feature plots, Density plots, and vlnplots genes of interest

if(long.compute){
  
  
  # Gene expression on UMAP
  
  for(i in seq_along(goi)){
    
    # feature plot
    print(
      FeaturePlot_scCustom(nk.seurat, 
                           goi[i], 
                           pt.size = 0.5)
      
    )
    
    dev.copy(pdf, paste0(output.dir, "FeaturePlot_", goi[i], ".pdf"))
    dev.off()
    
    
    # density plot
    print(
      Plot_Density_Custom(nk.seurat, 
                          goi[i], 
                          custom_palette = batlow.pal)
    )
    dev.copy(pdf, paste0(output.dir, "Densityplot_", goi[i], ".pdf"))
    dev.off()
    
    
    
    
    # Plot signature across categories
    meta.data.vars <- c("meta_histology", "tissue", "Majortype", "celltype")
    
    for(j in seq_along(meta.data.vars)){
      print(
        VlnPlot_scCustom(nk.seurat,
                         goi[i], 
                         group.by = meta.data.vars[j],
                         pt.size = 0.1) + NoLegend()
      )
      dev.copy(pdf, paste0(output.dir, "VlnPlot_", goi[i], "_", meta.data.vars[j], ".pdf"))
      dev.off()
      
    }
  }
}


# write plotting function 

create_barplot <- function(contingency_table, plot_title, return_data = FALSE) {
  # Calculate percentages
  percentage_table <- sweep(contingency_table, 2, colSums(contingency_table), FUN = "/") * 100
  
  # Convert the result to a data frame
  percentage_df <- as.data.frame(percentage_table)
  
  # Filter rows where 'Var1' is TRUE
  filtered_df <- percentage_df %>%
    filter(Var1 == TRUE)
  
  # Calculate the average frequency
  average_freq <- mean(filtered_df$Freq)
  
  # Load the ggplot2 library
  library(ggplot2)
  
  # Create the barplot with ordered data
  plot <- ggplot(filtered_df, aes(x = reorder(Var2, -Freq), y = Freq)) +
    geom_col(fill = "#112446") +
    geom_hline(yintercept = average_freq, linetype = "dotted", color = "red") +  # Add a red dotted line
    labs(x = "Category", 
         y = "Frequency of each category", 
         title = plot_title) +  # Use the input plot_title
    coord_flip() +
    theme_minimal() +
    theme(plot.title = element_text(hjust = 0.5))
  
  if (return_data) {
    return(filtered_df)
  } else {
    print(plot)
    return(NULL)
  }
}




# extract statistics and create barplots of score positive cell frequencies 

if(long.compute){
  
  for(i in seq_along(goi)){
    
    
    
    goi.expression <- nk.seurat@assays$RNA@data[goi[i], ]
    goi.logic <- goi.expression >= 0.5
    
    nk.seurat@meta.data$goistatus <- goi.logic
    
    
    
    meta.data.vars <- c("meta_histology", "tissue", "Majortype", "celltype")
    for(j in seq_along(meta.data.vars)){
      
      
      # Create a contingency table of 'sigstatus' and 'celltype'
      contingency_table <- table(nk.seurat@meta.data$goistatus, 
                                 nk.seurat@meta.data[[meta.data.vars[j]]])
      
      print(
        create_barplot(contingency_table, meta.data.vars[j])
      )
      
      dev.copy(pdf, paste0(output.dir, "Barplot_freq_of_cells_positive_for_", goi[i], "_across_", meta.data.vars[j], ".pdf"))
      dev.off()
      
      
      # export data
      output.dataframe <- create_barplot(contingency_table, meta.data.vars[j], return_data = TRUE)
      
      write.csv(output.dataframe, paste0(output.dir, "Freq_of_cells_positive_for_", goi[i], "_across_", meta.data.vars[j], ".csv"))
      
      
    }
    
    
  }
}


```

### Heatmap, vlnplot, densityplot, featureplot and % expression of gene sets of interest
```{r gene_sets_of_interest_analysis}

# Load necessary libraries
library(Seurat)
library(ggplot2)
library(stringr)
library(dplyr)



# Create barplot function
create_barplot <- function(contingency_table, plot_title, return_data = FALSE) {
  percentage_table <- sweep(contingency_table, 2, colSums(contingency_table), FUN = "/") * 100
  percentage_df <- as.data.frame(percentage_table)
  filtered_df <- percentage_df %>% filter(Var1 == TRUE)
  average_freq <- mean(filtered_df$Freq)
  
  plot <- ggplot(filtered_df, aes(x = reorder(Var2, -Freq), y = Freq)) +
    geom_col(fill = "#112446") +
    geom_hline(yintercept = average_freq, linetype = "dotted", color = "red") +
    labs(x = "Category", y = "Frequency of each category", title = plot_title) +
    coord_flip() +
    theme_minimal() +
    theme(plot.title = element_text(hjust = 0.5))
  
  if(return_data){
    return(filtered_df)
  } else {
    print(plot)
    return(NULL)
  }
}

# Define input path for gene lists
input.path <- "data/susanna_signatures/"
gene.list.files <- list.files(input.path, pattern = "\\.txt$", full.names = TRUE)

gene.sets <- lapply(gene.list.files, function(f){
  df <- read.table(f, header = TRUE, sep = "\t", stringsAsFactors = FALSE)
  return(as.character(df[[1]]))
})

names(gene.sets) <- basename(gene.list.files) |>
  str_replace_all(" ", "_") |>
  str_remove("\\.txt$")

# Set default assay
DefaultAssay(nk.seurat) <- "RNA"


# create output dir 
for(gene.set.name in names(gene.sets)){
  
  output.dir <- paste0("results/susanna_proteomics/", gene.set.name, "/")
  
  if (!dir.exists(output.dir)) {
    dir.create(output.dir, recursive = TRUE)
  }
  
}



clusters.var <- c("Majortype", "meta_tissue")

# heatmaps
for(cluster.var in clusters.var){
  
  # Compute pseudobulk once per cluster var
  pseudo.nk <- AggregateExpression(
    nk.seurat,
    assays        = "RNA",
    return.seurat = TRUE,
    group.by      = cluster.var
  )
  
  Idents(pseudo.nk) <- cluster.var
  
  cluster_genes <- function(seurat_obj, gene_list) {
    valid_genes <- intersect(gene_list, rownames(GetAssayData(seurat_obj, layer = "data")))
    if(length(valid_genes) == 0){
      warning("No valid genes found for clustering in ", cluster.var)
      return(gene_list)
    }
    expression_matrix <- GetAssayData(seurat_obj, layer = "data")[valid_genes, , drop = FALSE]
    dist_matrix <- dist(expression_matrix, method = "euclidean")
    hc <- hclust(dist_matrix, method = "ward.D2")
    ordered_genes <- valid_genes[hc$order]
    return(ordered_genes)
  }
  
  for(gene.set.name in names(gene.sets)){
    
    genes <- gene.sets[[gene.set.name]]
    ordered_genes <- cluster_genes(pseudo.nk, genes)
    
    output.dir <- paste0("results/susanna_proteomics/", gene.set.name, "/")
    
    if(length(ordered_genes) > 0){
      heat <- DoHeatmap(
        pseudo.nk,
        features    = ordered_genes,
        draw.lines  = FALSE,
        raster      = FALSE
      ) +
        scale_fill_gradient2(low = "blue", mid = "white", high = "red") +
        ggtitle(paste("Pseudobulk heatmap:", gene.set.name))
      
      ggsave(
        filename = paste0(output.dir, "Heatmap_", gene.set.name, "_", cluster.var, ".pdf"),
        plot     = heat, width = 8, height = 10
      )
    } else {
      warning("Skipping heatmap for ", gene.set.name, " in ", cluster.var, " (no valid genes).")
    }
    
  }
}


# now for each gene set, show vlnplot, featureplot, densityplot and % of expression above 0.5
for(gene.set.name in names(gene.sets)){
  
  genes <- gene.sets[[gene.set.name]]
  output.dir <- paste0("results/susanna_proteomics/", gene.set.name, "/")
  
  for(gene in genes){
    
    # check if gene can be found, otherwise skip
    valid_genes <- intersect(gene, rownames(GetAssayData(nk.seurat, layer = "data")))
    if(length(valid_genes) == 0){
      warning("Gene ", gene, " Not found, skipping to next gene")
      next()
    }



    # FeaturePlot
    fp <- FeaturePlot_scCustom(nk.seurat, gene, pt.size = 0.5)
    ggsave(paste0(output.dir, "FeaturePlot_", gene, ".pdf"), fp)
    
    # Density plot
    dp <- Plot_Density_Custom(nk.seurat, gene, custom_palette = batlow.pal)
    ggsave(paste0(output.dir, "Densityplot_", gene, ".pdf"), dp)
    
    # VlnPlots
    meta.data.vars <- c("meta_histology", "tissue", "Majortype", "celltype")
    for(var in meta.data.vars){
      vp <- VlnPlot_scCustom(nk.seurat, gene, group.by = var, pt.size = 0.1) + NoLegend()
      ggsave(paste0(output.dir, "VlnPlot_", gene, "_", var, ".pdf"), vp)
    }
    
    # Frequency-based barplots
    goi.expression <- nk.seurat@assays$RNA@data[gene, ]
    goi.logic <- goi.expression >= 0.5
    nk.seurat@meta.data$goistatus <- goi.logic
    
    for(var in meta.data.vars){
      contingency_table <- table(nk.seurat@meta.data$goistatus, nk.seurat@meta.data[[var]])
      barplot.obj <- create_barplot(contingency_table, var)
      dev.copy(pdf, paste0(output.dir, "Barplot_freq_of_cells_positive_for_", gene, "_across_", var, ".pdf"))
      dev.off()
      
      output.dataframe <- create_barplot(contingency_table, var, return_data = TRUE)
      write.csv(output.dataframe, paste0(output.dir, "Freq_of_cells_positive_for_", gene, "_across_", var, ".csv"))
    }
  }
}




```

### Frequency of expression plots aggregated by sample and with stats 
```{r freq_expression_plots}
# Load necessary libraries
library(Seurat)
library(ggplot2)
library(stringr)
library(dplyr)
library(ggpubr)

# Create barplot function with ANOVA and posthoc
plot_gene_expression_proportion <- function(seurat_obj, gene, group.var, threshold.val, out.dir, min.group.size = 0, y.axis.max = 110, xlabel.rotation = 90) {
  expr.values <- GetAssayData(seurat_obj, assay = "RNA", slot = "data")[gene, ]
  seurat_obj@meta.data[[paste0(gene, "_positive")]] <- expr.values >= threshold.val

  plot_data <- seurat_obj@meta.data %>%
    mutate(donor_id = .data$meta_patientID) %>%
    group_by(donor_id, group = .data[[group.var]], gene_status = .data[[paste0(gene, "_positive")]]) %>%
    summarise(freq = n(), .groups = "drop_last") %>%
    ungroup() %>%
    group_by(donor_id, group) %>%
    mutate(prop = freq / sum(freq) * 100) %>%
    filter(gene_status == TRUE)

  # Filter groups by minimum donor size if required
  if(min.group.size > 0){
    group_counts <- plot_data %>% group_by(group) %>% summarise(n = n_distinct(donor_id))
    valid_groups <- group_counts$group[group_counts$n >= min.group.size]
    plot_data <- plot_data %>% filter(group %in% valid_groups)
  }

  # Ensure 'group' is a character/factor for plotting
  plot_data$group <- as.factor(plot_data$group)

  # Export raw calculated values
  write.table(plot_data, file = paste0(out.dir, gene, "_", group.var, "_calculated_frequencies.txt"), sep = "\t", row.names = FALSE, quote = FALSE)

  # Save ANOVA and posthoc stats
  stats.outfile <- paste0(out.dir, gene, "_", group.var, "_ANOVA_posthoc.txt")
  sink(stats.outfile)
  aov.res <- aov(prop ~ group, data = plot_data)
  print(summary(aov.res))
  posthoc <- TukeyHSD(aov.res)
  print(posthoc)
  sink()

  # Group combinations for t-tests
  comparisons <- combn(levels(plot_data$group), 2, simplify = FALSE)

  # n values
  legend_data <- plot_data %>% group_by(group) %>% summarise(n = n_distinct(donor_id))
  group.levels <- levels(plot_data$group)

  plot <- ggplot(plot_data, aes(x = group, y = prop, color = group, fill = group)) +
    geom_boxplot(alpha = 0.3, outlier.shape = NA, width = 0.75) +
    geom_point(position = position_jitter(width = 0.1), shape = 19, size = 2) +
    labs(x = group.var, y = paste0("% of ", gene, " positive cells")) +
    theme_minimal() +
    stat_compare_means(method = "anova", label.y = y.axis.max - 5) +
    stat_compare_means(comparisons = comparisons, method = "t.test", label = "p.signif", hide.ns = FALSE) +
    scale_y_continuous(expand = expansion(add = c(0, 1)), limits = c(0, y.axis.max)) +
    scale_colour_manual(values = seq_along(group.levels),
                        labels = paste(group.levels, " (n=", legend_data$n, ")", sep = ""),
                        aesthetics = c("colour", "fill")) +
    theme(axis.text.x = element_text(angle = xlabel.rotation, vjust = 0.5, hjust = 1)) +
    ggtitle(paste("Proportion of", gene, "positive cells by", group.var))


  ggsave(filename = paste0(out.dir, gene, "_", group.var, "_barplot.pdf"),
         plot = plot, width = 7, height = 5)
}

# Define threshold and output directory
threshold.val <- 0.5
output.dir <- "results/susanna_dgat_quantification/"
dir.create(output.dir, showWarnings = FALSE, recursive = TRUE)

# Genes of interest
goi <- c("DGAT1")

# Analysis across tissue
for(gene in goi){
  plot_gene_expression_proportion(
    seurat_obj  = nk.seurat,
    gene        = gene,
    group.var   = "tissue",
    threshold.val = threshold.val,
    out.dir     = output.dir,
    xlabel.rotation = 90,
    y.axis.max  = 100
  )
}

# Analysis across Majortype
for(gene in goi){
  plot_gene_expression_proportion(
    seurat_obj  = nk.seurat,
    gene        = gene,
    group.var   = "Majortype",
    threshold.val = threshold.val,
    out.dir     = output.dir,
    xlabel.rotation = 90,
    y.axis.max  = 100
  )
}

# Analysis across meta_histology with filtering for min 10 donors
for(gene in goi){
  plot_gene_expression_proportion(
    seurat_obj  = nk.seurat,
    gene        = gene,
    group.var   = "meta_histology",
    threshold.val = threshold.val,
    out.dir     = output.dir,
    xlabel.rotation = 90,
    min.group.size = 10,
    y.axis.max  = 100
  )
}


# note in output 
#freq: number of cells in each donor–group combination that are either positive or negative for the gene.
#prop: proportion (in %) of gene-positive cells in that group for each donor.
```







### Heatmaps genes of interest with row/gene clustering
```{r goi_heatmaps_gene_clustered}

# Set default assay
DefaultAssay(nk.seurat) <- "RNA"

clusters.var <- c("Majortype", "meta_tissue")

for(cluster.var in clusters.var){
  
  # Compute pseudobulk once per cluster var
  pseudo.nk <- AggregateExpression(
    nk.seurat,
    assays        = "RNA",
    return.seurat = TRUE,
    group.by      = cluster.var
  )
  
  Idents(pseudo.nk) <- cluster.var
  
  cluster_genes <- function(seurat_obj, gene_list) {
    valid_genes <- intersect(gene_list, rownames(GetAssayData(seurat_obj, layer = "data")))
    if(length(valid_genes) == 0){
      warning("No valid genes found for clustering in ", cluster.var)
      return(gene_list)
    }
    expression_matrix <- GetAssayData(seurat_obj, layer = "data")[valid_genes, , drop = FALSE]
    dist_matrix <- dist(expression_matrix, method = "euclidean")
    hc <- hclust(dist_matrix, method = "ward.D2")
    ordered_genes <- valid_genes[hc$order]
    return(ordered_genes)
  }
  
  for(gene.set.name in names(gene.sets)){
    
    genes <- gene.sets[[gene.set.name]]
    ordered_genes <- cluster_genes(pseudo.nk, genes)
    
    if(length(ordered_genes) > 0){
      heat <- DoHeatmap(
        pseudo.nk,
        features    = ordered_genes,
        draw.lines  = FALSE,
        raster      = FALSE
      ) +
        scale_fill_gradient2(low = "blue", mid = "white", high = "red") +
        ggtitle(paste("Pseudobulk heatmap:", gene.set.name))
      
      ggsave(
        filename = paste0(output.dir, "Heatmap_", gene.set.name, "_", cluster.var, ".pdf"),
        plot     = heat, width = 8, height = 10
      )
    } else {
      warning("Skipping heatmap for ", gene.set.name, " in ", cluster.var, " (no valid genes).")
    }
  }
}





```


### heatmaps genes of interest
```{r DEG_heatmaps}

clusters.var <- c("Majortype", "meta_tissue")  # Annotation levels

for (cluster.var in clusters.var) {
  
  # Replace "." with "_" in cluster.var for filenames and directories
  cluster.var.clean <- gsub("\\.", "_", cluster.var)
  
  # Create output directory dynamically
  output.dir <- paste0("results/Susanna_Proteomics_Signatures_", cluster.var.clean, "/")
  
  if (!dir.exists(output.dir)) {
    dir.create(output.dir, recursive = TRUE)
  }
  
  DefaultAssay(nk.seurat) <- "RNA"
  
  # Create pseudobulk for visualization
  pseudo.nk <- AggregateExpression(nk.seurat, 
                                   assays = "RNA",
                                   return.seurat = TRUE,
                                   group.by = cluster.var)
  
  # Set idents of pseudobulk object
  Idents(pseudo.nk) <- cluster.var
  
  # Function to cluster genes using Ward's D2 and Euclidean distance
  cluster_genes <- function(seurat_obj, gene_list) {
    # Ensure only genes that exist in the dataset are used
    valid_genes <- intersect(gene_list, rownames(GetAssayData(seurat_obj, layer = "data")))
    
    if (length(valid_genes) == 0) {
      warning("No valid genes found for clustering in ", cluster.var)
      return(gene_list)  # Return original order if no valid genes found
    }
    
    # Extract expression matrix for selected genes
    expression_matrix <- GetAssayData(seurat_obj, layer = "data")[valid_genes, , drop = FALSE]
    
    # Compute distance matrix (Euclidean)
    dist_matrix <- dist(expression_matrix, method = "euclidean")
    
    # Perform hierarchical clustering using Ward.D2
    hc <- hclust(dist_matrix, method = "ward.D2")
    
    # Order genes by clustering result
    ordered_genes <- valid_genes[hc$order]
    
    return(ordered_genes)
  }
  
  # Cluster genes for heatmaps (handle missing genes)
  ordered_vehicle_genes <- cluster_genes(pseudo.nk, DE.in.vehicle.only)
  ordered_inhibitor_genes <- cluster_genes(pseudo.nk, DE.in.inhibitor.only)
  
  # Heatmap of vehicle DEG signature (with clustering)
  if (length(ordered_vehicle_genes) > 0) {
    print(
      DoHeatmap(pseudo.nk, 
                features = ordered_vehicle_genes,  # Clustered gene order
                draw.lines = FALSE, 
                raster = FALSE)
    )
    
    dev.copy(pdf, paste0(output.dir, "Heatmap_vehicle_DEG_signature_", cluster.var.clean, "_pseudobulk.pdf"))
    dev.off()
  } else {
    warning("Skipping heatmap for vehicle genes in ", cluster.var, " (no valid genes).")
  }
  
  # Heatmap of inhibitor DEG signature (with clustering)
  if (length(ordered_inhibitor_genes) > 0) {
    print(
      DoHeatmap(pseudo.nk, 
                features = ordered_inhibitor_genes,  # Clustered gene order
                draw.lines = FALSE, 
                raster = FALSE)
    )
    
    dev.copy(pdf, paste0(output.dir, "Heatmap_inhibitor_DEG_signature_", cluster.var.clean, "_pseudobulk.pdf"))
    dev.off()
  } else {
    warning("Skipping heatmap for inhibitor genes in ", cluster.var, " (no valid genes).")
  }
}

```
